Estimating the Financial Impact of Wildfire Smoke on US National parks

### 2020-05-27 Markdown Documents/Manuscript.Rmd
added 2 paragraphs to the intro

---
title: Estimating the Financial Impact of Wildfire Smoke on US National parks
author:
- affiliation: Boise State University - Human Environment Systems
  name: Matt Clark
- affiliation: University of Michigan
  name: Alexander Killion
- affiliation: Boise State University
  name: Matthew A. Williamson
- affiliation: Boise State University
  name: Vicken?
- affiliation: Boise State University
  name: Laura? (undergrad who will continue working over summer)
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    df_print: paged
  word_document: default
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
    template: /Users/Matthewclark989/Documents/SmokeProject/svm_ms_template.tex
fontfamily: mathpazo
fontsize: 11pt
geometry: margin=1in
keywords: pandoc, r markdown, knitr
bibliography: /Users/Matthewclark989/Documents/SmokeProject/References.bib
thanks: If we want to thank anyone? Maybe Moji?.
abstract: The United States national park system serves as a global model for how natural resource protection can provide numerous social and ecological services while generating economic returns on investment. In recent years parks have increasingly relied on visitation fees to fund consevation projects as well as basic infrastructure. In contrast with congressional funding, visitation is subject to fluctuation given climatic and environmental variability. While many environmental variables are relatively predictable and stable over time, wildfire smoke events have increased drmatically via climate change, particurlarly in the American West. In this project, we assess how recreational visitation to national parks has systematically varied with wildfire smoke. Over the last 38 years, we estimate that wildfire smoke has led to a 12% overall visitation decrease to National Parks in the Pacific West and Inter-Mountain Regions. When accounting for visitation fees, this loss may have cost the National Park Service as much as $580 million in total. Wildfire frequency and severity are expected to increase considerably in these regions over the next 50 years. Assuming current trends in smoke, park visitation, and park fees continue, we expect to see significant financial losses to the National Park Service via wildfire smoke in the coming years. Our recomendation is that....NPS work with the forest service to prevent fires in areas close to parks?? Do they already do that??    
---
# 1. Introduction

@@

As national parks become increasingly dependent on visitor fees in order to provide the suite of social and ecological services communities depend on, it is of paralleled importance to understand what drives fluctuation in recreational visits. Research has shown that economic considerations (e.g. park fees) are considerably less influential on park visitation than are recreational opportunities [@neuvonen_visits_2010;@stevens_declining_2014]. Most factors which influence the recreational accessibility of a given park are either fixed (natural resources within the park area) or within the control of park management (permitting, trail building, etc.). Wildfire smoke however, shows considerable yearly variation and is largely produced by adjacent public lands and therefore out of the jurisdiction of park management [@dombeck_wildfire_2004;@collins_regional_2006].

Wildfire smoke is a mixture of many gasses aerated by biomass combustion across a landscape [@naeher_woodsmoke_2007]. While many of these compounds have been shown to cause health problems in humans, organic carbon and black carbon particulate matter less than 2.5 μm in diameter (PM2.5) is particularly harmful to human health, causing between 260,000 - 600,000 global deaths annually [@johnston_fay_h_estimated_2012; @naeher_woodsmoke_2007]. Health concerns regarding wildfire smoke, specifically PM2.5, regularly drive calls for individuals to reduce outdoor recreation behavior [@oneill_summary_2013].

Wildfires and the smoke they produce have been increasing in the United States over the last 50 years, specifically in the American West and Inter-Mountain regions [@kitzberger_contingent_2007;@westerling_increasing_2016]. These regions are also projected to experience increased wildfire frequency and severity over the next 50 years as a result of climate change [@harvey_human-caused_2016; @peterson_new_2003; @barbero_climate_2015]. Anecdotally, it is known that wildfire smoke has impacted the recreation behavior of individuals in these areas and decreased visitation revenue to the national parks in both regions. This assumption however has not yet been demonstrated empirically.

The primary goal of this project is to assess how visitation to national parks in the Inter-Mountain and Pacific regions has systematically varied with surface level concentrations of PM2.5 resulting from wildfire smoke from 1980 - 2018. Given the variability in recreational fees between parks and over time, the secondary goal of this project is to produce order of magnitude estimates of the financial losses that wildfire smoke has caused  park units during the same period. The intention of this research is not to perform an economic analysis of smoke effects on each park specifically, but rather to provide an approximation of how wildfire smoke affects population level recreational decision making and what the downstream economic impacts have been historically and may be in the future.

>>>>

---
title: Estimating the Financial Impact of Wildfire Smoke on US National parks
author:
- affiliation: Boise State University - Human Environment Systems
  name: Matt Clark
- affiliation: University of Michigan
  name: Alexander Killion
- affiliation: Boise State University
  name: Matthew A. Williamson
- affiliation: Boise State University
  name: Vicken?
- affiliation: Boise State University
  name: Laura? (undergrad who will continue working over summer)
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    df_print: paged
  word_document: default
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
    template: /Users/Matthewclark989/Documents/SmokeProject/svm_ms_template.tex
fontfamily: mathpazo
fontsize: 11pt
geometry: margin=1in
keywords: pandoc, r markdown, knitr
bibliography: /Users/Matthewclark989/Documents/SmokeProject/References.bib
thanks: If we want to thank anyone? Maybe Moji?.
abstract: The United States national park system serves as a global model for how natural resource protection can provide numerous social and ecological services while generating economic returns on investment. In recent years parks have increasingly relied on visitation fees to fund consevation projects as well as basic infrastructure. In contrast with congressional funding, visitation is subject to fluctuation given climatic and environmental variability. While many environmental variables are relatively predictable and stable over time, wildfire smoke events have increased drmatically via climate change, particurlarly in the American West. In this project, we assess how recreational visitation to national parks has systematically varied with wildfire smoke. Over the last 38 years, we estimate that wildfire smoke has led to a 12% overall visitation decrease to National Parks in the Pacific West and Inter-Mountain Regions. When accounting for visitation fees, this loss may have cost the National Park Service as much as $580 million in total. Wildfire frequency and severity are expected to increase considerably in these regions over the next 50 years. Assuming current trends in smoke, park visitation, and park fees continue, we expect to see significant financial losses to the National Park Service via wildfire smoke in the coming years.     
---
# 1. Introduction

@@

As national parks become increasingly dependent on visitor fees in order to provide the suite of social and ecological services communities depend on, it is of paralleled importance to understand what drives fluctuation in recreational visits. Research has shown that economic considerations (e.g. park fees) are considerably less influential on park visitation than are recreational opportunities [@neuvonen_visits_2010;@stevens_declining_2014]. Most factors which influence the recreational accessibility of a given park are either fixed (e.g. natural resources within the park area) or within the control of park management (permitting, trail building, etc.). Wildfires however, largely result from external forces and have been shown to negatively affect individual's willingness to engage in recreational travel [@thapa_wildfires_2013]. Furthermore, a study of five national parks in Utah showed visitation losses from wildfire events even at a distance of up to 320 kilometers from the park boundary [@kim_wildfire_2019]. Given the distance with which wildfires have been shown to impact park visitation, it can be assumed that individuals modify their recreational behavior not only as a result of wildfires themselves but also the smoke they produce.

Wildfire smoke is a mixture of many gasses aerated by biomass combustion across a landscape [@naeher_woodsmoke_2007]. While many of these compounds have been shown to cause health problems in humans, organic carbon and black carbon particulate matter less than 2.5 μm in diameter (PM2.5) is particularly harmful to human health, causing between 260,000 - 600,000 global deaths annually [@johnston_fay_h_estimated_2012; @naeher_woodsmoke_2007]. A 2012 study estimated that wildfire smoke in Los Angeles County resulted in public health costs of approximately $84 per exposed person each day [@richardson_hidden_2012]. Health concerns regarding wildfire smoke, specifically PM2.5, regularly drive calls for individuals to reduce outdoor recreational behavior [@oneill_summary_2013].

Recent literature reviews on recreational behavior have found a dearth of research both describing how individual's are affected by and respond to low air quality, as well as research which gives predictions for how climate change will affect the recreation landscape in specific areas [@otoole_climate_2019;@zajchowski_air_2019-1]. Additionally, a survey of federal land managers indicated that decision makers in areas such as national parks feel that they have a shortage of information describing how air quality affects the recreational behavior of their visitors [@zajchowski_air_2019]. It is of particular importance to understand how individuals respond to wildfire smoke because of the highly variable nature of wildfires, their relationship to climate change, and the great distances that smoke can travel during and after wildfire events [@dombeck_wildfire_2004; @collins_regional_2006].

Wildfires and the smoke they produce have been increasing in the United States over the last 50 years, specifically in the American West and Inter-Mountain regions [@kitzberger_contingent_2007;@westerling_increasing_2016]. These regions are also projected to experience increased wildfire frequency and severity over the next 50 years as a result of climate change [@harvey_human-caused_2016; @peterson_new_2003; @barbero_climate_2015].

The primary goal of this project is to assess how visitation to national parks in the Inter-Mountain and Pacific West regions has systematically varied with surface level concentrations of PM2.5 resulting from wildfire smoke from 1980 - 2018. Given the variability in recreational fees between parks and over time, the secondary goal of this project is to produce order of magnitude estimates of the financial losses that wildfire smoke has caused park units during the same period. The intention of this research is not to perform an in-depth economic analysis of smoke effects on each park specifically, but rather to provide an approximation of how wildfire smoke affects population level recreational decision making and what the downstream economic impacts have been historically and what the implications are for the future. Overall, this research aims to better equip protected area managers with the information necessary to best manage recreational resources in the face of increasing wildfire events and climatic variability generally.
